Ore Polynomials in Sage\thanks{All three authors were supported by the Austrian FWF grant Y464-N18.}

### 2014-02-27 paper/main.tex

\documentclass[11pt]{amsart}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bbm}
\usepackage{a4wide}
\usepackage{sagetex}
\usepackage{tikz}

@@

 \author[Manuel Kauers, Maximilian Jaroschek, Fredrik Johansson]
   {Manuel Kauers\,$^\ast$, Maximilian Jaroschek\,$^\ast$, Fredrik Johansson\,$^\ast$}
 \address{Manuel Kauers, Research Institute for Symbolic Computation (RISC), J. Kepler University Linz, Austria}
 \email{mkauers@risc.uni-linz.ac.at}
 \address{Maximilian Jaroschek, Research Institute for Symbolic Computation (RISC), J. Kepler University Linz, Austria}
 \email{mjarosch@risc.uni-linz.ac.at}
 \address{Fredrik Johansson, Research Institute for Symbolic Computation (RISC), J. Kepler University Linz, Austria}
 \email{fjohanss@risc.uni-linz.ac.at}
 \thanks{$^\ast$ Supported by the Austrian FWF grant Y464-N18.}

 \title{Ore Polynomials in Sage}

 \begin{abstract}
We present a Sage implementation of Ore algebras. The main features for the most
common instances include basic arithmetic and actions; gcrd and lclm; D-finite
closure properties; natural transformations between related algebras; guessing;
desingularization; solvers for polynomials, rational functions and (generalized)
power series. This paper is a tutorial on how to use the package.
 \end{abstract}

@@

@@

Algorithms for Ore algebras provide a systematic answer to this
question~\cite{bronstein96,chyzak98}.  Invented in the first half of the 20th
century~\cite{ore33} with the objective of providing a unified theory for
various kinds of linear operators, they have been used for many years in
computer algebra systems, for example in the Maple packages
OreTools~\cite{abramov03}, gfun~\cite{salvy94} or mgfun~\cite{chyzak98a}, or in
the Mathematica packages by Mallinger~\cite{mallinger96} and
Koutschan~\cite{koutschan10c,koutschan09}.

@@

For Ore algebras with several generators, it is possible to mix specifications
of generators via triples $(X,\sigma,\delta)$ with generators using the naming
convention shortcuts as explained before. Continuing the previous example,
here is a way to define an algebra $A$ over $K$ with two generators, a $D$ 
that behaves like before, and in addition an $Sx$ which acts like the standard
shift on~$x$ and leaves the other generators fixed.

@@

The efficiency of computing the GCRD depends on the size of the
coefficients of intermediate results, and there are different strategies to
control this growth via so called polynomial remainder sequences (PRS). The default is
the improved PRS described in~\cite{jaroschek13b}, which will usually be the
fastest choice. Other strategies can be selected by the option \verb|prs|.

@@

If $L_1,L_2$ are operators, then the solutions of their GCRD are precisely the
common solutions of $L_1$ and~$L_2$. The LCLM, on the other hand, is the minimal
order operator whose solution space contains all the solutions of $L_1$ and
all the solutions of~$L_1$. Because of this property, GCRD and LCLM are
useful tools for constructing operators with prescribed solutions. For example,
here is how to construct a differential operator which has the solutions $x^5$
and $\exp(x)$, starting from the obvious operators annihilating $x^5$
and~$\exp(x)$, respectively.

@@

Guessing is, in some sense, the reverse operation of \verb|to_list| for
recurrence operators, or of \verb|power_series_solutions| for differential
operators.  It is of the most popular features of packages like gfun, and there
are even some special purpose packages dedicated to this
technique~\cite{kauers09a,hebisch11}. The basic idea is simple. Given
a finite array of numbers, thought of as the first terms of an infinite
sequence, we want to know whether this sequence satisfies a recurrence. The
algorithm behind a guessing engine searches for small equations matching the
given data.  Generically, no such equations exist, so if some are found, it is
fair to ``guess'' that they are in fact valid equations for the whole infinite
sequence.

@@

\bibitem{abramov03}
Sergei~A. Abramov, Ha~Q. Le, and Ziming Li.
\newblock {OreTools:} a computer algebra library for univariate ore polynomial
  rings.
\newblock Technical Report CS-2003-12, University of Waterloo, 2003.

@@

\bibitem{jaroschek13b}
Maximilian Jaroschek.
\newblock Improved polynomial remainder sequences for {Ore} polynomials.
\newblock {\em Journal of Symbolic Computation}, 2013.
\newblock to appear.

\bibitem{jaroschek13a}
Maximilian Jaroschek, Manuel Kauers, Shaoshi Chen, and Michael~F. Singer.
\newblock Desingularization explains order-degree curves for {Ore} operators.
\newblock In Manuel Kauers, editor, {\em Proceedings of ISSAC'13}, 2013.
\newblock to appear.

@@

\bibitem{kauers13}
Manuel Kauers.
\newblock The holonomic toolkit.
\newblock In Johannes Bluemlein and Carsten Schneider, editors, {\em Computer
  Algebra in Quantum Field Theory: Integration, Summation and Special
  Functions}, Texts and Monographs in Symbolic Computation. Springer, 2013.

>>>>

%\documentclass[11pt]{amsart}
\documentclass{llncs}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bbm}
\usepackage{a4wide}
\usepackage{sagetex}
\usepackage{tikz}

@@

 \title{Ore Polynomials in Sage\thanks{All three authors were supported by the Austrian FWF grant Y464-N18.}}
 \titlerunning{Ore Polynomials in Sage}

 \author{Manuel Kauers\and Maximilian Jaroschek\and Fredrik Johansson}
 \authorrunning{Manuel Kauers\and Maximilian Jaroschek\and Fredrik Johansson}

\institute{Research Institute for Symbolic Computation (RISC)\\ 
  Johannes Kepler University (JKU)\\
  4040 Linz, Austria\\
  \email{[mkauers|mjarosch|fjohanss]@risc.jku.at}}

@@

 \begin{abstract}
We present a Sage implementation of Ore algebras. The main features for the most
common instances include basic arithmetic and actions; GCRD and LCLM; D-finite
closure properties; natural transformations between related algebras; guessing;
desingularization; solvers for polynomials, rational functions and (generalized)
power series. This paper is a tutorial on how to use the package.
 \end{abstract}

@@

Algorithms for Ore algebras provide a systematic answer to this
question~\cite{bronstein96,chyzak98}.  Invented in the first half of the 20th
century~\cite{ore33} with the objective of providing a unified theory for
various kinds of linear operators, they have been used for many years in
computer algebra systems, for example in the Maple packages
OreTools~\cite{abramov03}, gfun~\cite{salvy94} or Mgfun~\cite{chyzak98a}, or in
the Mathematica packages by Mallinger~\cite{mallinger96} and
Koutschan~\cite{koutschan10c,koutschan09}.

@@

For Ore algebras with several generators, it is possible to mix specifications
of generators via triples $(X,\sigma,\delta)$ with generators using the naming
convention shortcuts as explained before. Continuing the previous example,
here is a way to define an algebra $A$ over $K$ with two generators, a $D$ 
that behaves like before, and in addition an $\mathrm{Sx}$ which acts like the standard
shift on~$x$ and leaves the other generators fixed.

@@

The efficiency of computing the GCRD depends on the size of the
coefficients of intermediate results, and there are different strategies to
control this growth via so-called polynomial remainder sequences (PRS). The default is
the improved PRS described in~\cite{jaroschek13b}, which will usually be the
fastest choice. Other strategies can be selected by the option \verb|prs|.

@@

If $L_1,L_2$ are operators, then the solutions of their GCRD are precisely the
common solutions of $L_1$ and~$L_2$. The LCLM, on the other hand, is the minimal
order operator whose solution space contains all the solutions of $L_1$ and
all the solutions of~$L_2$. Because of this property, GCRD and LCLM are
useful tools for constructing operators with prescribed solutions. For example,
here is how to construct a differential operator which has the solutions $x^5$
and $\exp(x)$, starting from the obvious operators annihilating $x^5$
and~$\exp(x)$, respectively.

@@

Guessing is, in some sense, the reverse operation of \verb|to_list| for
recurrence operators, or of \verb|power_series_solutions| for differential
operators.  It is one of the most popular features of packages like gfun, and there
are even some special purpose packages dedicated to this
technique~\cite{kauers09a,hebisch11}. The basic idea is simple. Given
a finite array of numbers, thought of as the first terms of an infinite
sequence, we want to know whether this sequence satisfies a recurrence. The
algorithm behind a guessing engine searches for small equations matching the
given data.  Generically, no such equations exist, so if some are found, it is
fair to ``guess'' that they are in fact valid equations for the whole infinite
sequence.

@@

\bibitem{abramov03}
Sergei~A. Abramov, Ha~Q. Le, and Ziming Li.
\newblock {OreTools:} a computer algebra library for univariate Ore polynomial
  rings.
\newblock Technical Report CS-2003-12, University of Waterloo, 2003.

@@

\bibitem{jaroschek13b}
Maximilian Jaroschek.
\newblock Improved polynomial remainder sequences for {Ore} polynomials.
\newblock {\em Journal of Symbolic Computation}, 58:64--76, 2013.

\bibitem{jaroschek13a}
Maximilian Jaroschek, Manuel Kauers, Shaoshi Chen, and Michael~F. Singer.
\newblock Desingularization explains order-degree curves for {Ore} operators.
\newblock In Manuel Kauers, editor, {\em Proceedings of ISSAC'13}, pp. 157--164, 2013.

@@

\bibitem{kauers13}
Manuel Kauers.
\newblock The holonomic toolkit.
\newblock In Johannes Bl\"umlein and Carsten Schneider, editors, {\em Computer
  Algebra in Quantum Field Theory: Integration, Summation and Special
  Functions}, Texts and Monographs in Symbolic Computation. Springer, 2013.
